---
title: "Analysis of the portfolios regarding the MSCI Spain"
author: "Emanuel Sommer"
output: 
  html_document:
    toc: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this analysis different portfolios consisting of stocks that are among the top 10 constituents of the MSCI Spain Index are backtested using the functionality of the `portvine` package. First the most important packages are loaded and after that the data will be imported and discussed shortly.

```{r, message=FALSE, warning=FALSE}
library(portvine)
library(tidyverse)
# utility color vector for visualizations
yes_no_cols <- c("#92B8DE", "#db4f59")
theme_set(
  theme_minimal() +
  theme(plot.title = ggtext::element_markdown(size = 11),
        plot.subtitle = ggtext::element_markdown(size = 9))
)
# load the data
load(here::here("data", "msci_spain_data_clean.RData"))
```

## A first glimpse at the data

```{r}
glimpse(msci_spain_data)
```

One can see that there are 11 columns. The `date` column gives obviously the date as daily return data will be analyzed here. The daily log returns of the overall index are given in the column `msci_spain_index` and all other columns contain the daily log returns of the respective stock. One can also detect missing values which can be nicely observed in the visualization below.

```{r, echo=FALSE}
msci_spain_data %>%
  rowid_to_column("id") %>%
  pivot_longer(-c(date, id)) %>%
  mutate(missing = is.na(value)) %>%
  ggplot() +
  geom_point(aes(x = date, y = factor(name), col = missing),
             shape = 15, size = 10) +
  scale_color_manual(values = yes_no_cols) +
  scale_y_discrete(labels = ~ str_to_title(str_replace_all(.x, "_", " "))) +
  guides(col = "none") +
  labs(x = "", y = "",
       title = paste("Display asset returns that are ",
                     "<span style='color:",
                     yes_no_cols[1],
                     "'>**available**</span>",
                     " or ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**missing**</span>",
                     "."))
  
```

This makes total sense as for example Cellnex Telecom stocks are traded since 2015, so there can not be earlier return data. So one could now either impute the values in order to use the whole time frame but here the more simplistic approach of just looking at a time frame without any missing values is applied.
This reduced data set is available as `msci_spain_complete_data` and contains only complete cases. 

```{r}
summary(msci_spain_complete_data$date)
nrow(msci_spain_complete_data)
```

This means that the time frame is from the 8th Mai of 2015 until the 8th November of 2021 and there are `r nrow(msci_spain_complete_data)` observations. Below one can have a look at daily log returns of the overall MSCI Spain index.

```{r, echo=FALSE}
ggplot(msci_spain_complete_data, aes(x = date, y = msci_spain_index)) +
  geom_line() +
  labs(x = "", y = "Daily log return",
       title = "MSCI Spain index") 
```

So one can observe greater volatility during the stock market selloffs 2015-2016 maybe due to Chinese stock market turbulences, the EU dept crisis and the Brexit votum as well as for the first 'Covid-19 year' 2020. This should give the opportunity to access how robust the risk measure estimates are during a crisis. Now specify the first portfolio and analyze it unconditionally as well as conditionally.


## The missing leader portfolio `missl`

Here one picks as the stock portfolio equally weighted all the stocks without the one with the highest market capitalization that makes roughly 17% of the MSCI Spain i.e. Iberdrola, that is mainly in the energy business. First one models this portfolio and estimates the risk estimates unconditionally and then conditionally on the biggest player Iberdrola. 

### Unconditional `missl` portfolio

First specify suitable marginal model settings as well as settings for the vine copula.

```{r}
missl_marginal_settings <- marginal_settings(
  train_size = 1000,
  refit_size = 50
)

missl_vine_settings <- vine_settings(
  train_size = 200,
  refit_size = 25,
  family_set = "all",
  vine_type = "rvine"
)
```

To compare run times one estimates the risk once sequentially and once in parallel.

```{r}
# sequentially
missl_risk_roll_seq <- estimate_risk_roll(
  msci_spain_complete_data %>% select(-c(date, msci_spain_index, iberdrola)),
  weights = NULL,
  marginal_settings = missl_marginal_settings,
  vine_settings = missl_vine_settings,
  alpha = c(0.01, 0.05, 0.1),
  risk_measures = c("VaR", "ES_mean", "ES_median", "ES_mc"),
  n_samples = 1000,
  n_mc_samples = 1000,
  trace = TRUE
)

# 4 parallel R session
future::plan("multisession", workers = 4)
missl_risk_roll <- estimate_risk_roll(
  msci_spain_complete_data %>% select(-c(date, msci_spain_index, iberdrola)),
  weights = NULL,
  marginal_settings = missl_marginal_settings,
  vine_settings = missl_vine_settings,
  alpha = c(0.01, 0.05, 0.1),
  risk_measures = c("VaR", "ES_mean", "ES_median", "ES_mc"),
  n_samples = 1000,
  n_mc_samples = 1000,
  trace = TRUE
)
future::plan("sequential")
```

The results and their respective run times can be seen below.

```{r}
missl_risk_roll_seq
missl_risk_roll
# thus the speedup is already for 4 workers
missl_risk_roll_seq@time_taken / missl_risk_roll@time_taken
# besides the print method one can also call the summary for more details
summary(missl_risk_roll)
```

#### Inspect marginal and vine copula model fits

Now one can use the accessor functions to extract the fitted marginal as well as the vine copula models.

```{r}
### get the fitted marginal models
missl_fitted_marginals <- fitted_marginals(missl_risk_roll)
# for each asset one now can access the uGARCHroll object that for example 
# contains the fitted coefficients for the second rolling window 
missl_fitted_marginals$banco_santander@model$coef[[2]]$coef

### get the fitted vine copula models
missl_fitted_vines <- fitted_vines(missl_risk_roll)
# then one can for example have a look at how the fitted vine structure evolved
#  over time by plotting the first tree for different rolloing windows
rvinecopulib:::plot.vinecop(missl_fitted_vines[[1]])
rvinecopulib:::plot.vinecop(missl_fitted_vines[[28]])
```

#### Analyze estimated risk measures

Now finally the risk estimates are going to be analyzed. First one compares the 3 different methods of estimating the Expected Shortfall.

```{r, echo=FALSE}
risk_estimates(missl_risk_roll,
               risk_measures = c("ES_mean", "ES_median", "ES_mc"),
               alpha = 0.05) %>%
  pivot_wider(names_from = risk_measure, values_from = risk_est) %>%
  plotly::plot_ly(x = ~ES_mean, y = ~ES_median, z = ~ES_mc,
                  marker = list(symbol = 'circle', color = "black",
                                size = 2),
                  text = ~ paste('Row:', row_num)) %>%
  plotly::add_markers(opacity = 0.2) %>%
  plotly::layout(
    scene = list(xaxis = list(title = 'Mean'),
                 yaxis = list(title = 'Median'),
                 zaxis = list(title = 'Monte Carlo')),
    title = "Empirical comparison of Expected Shortfall estimatiors"
  )
```

Also one might compare the number of exceedances.

```{r}
risk_estimates(missl_risk_roll,
               risk_measures = c("ES_mean", "ES_median", "ES_mc"),
               alpha = 0.05,
               exceeded = TRUE) %>%
  group_by(risk_measure) %>%
  summarise(relative_exceedances = mean(exceeded))
```

Here actually we find an indication for the fact that the values that fell below the corresponding VaR where left skewed which leads to the fact that the mean will be a more conservative estimate in those cases which might explain the slightly lower exceedance rate.

### Exceedance plots

```{r, echo=TRUE}
risk_estimates(missl_risk_roll,
               risk_measures = c("ES_mean", "VaR"),
               alpha = 0.01) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est,
                col = risk_measure, linetype = risk_measure),
            size = 0.5) +
  labs(x = "estimation window", y = "portfolio log returns",
       linetype = "Risk measure",
       title = "Comparison of risk measure behaivior for alpha level 0.01") +
  scale_color_manual(values = c(yes_no_cols[1], "#477042"),
                     name = "Risk measure") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

```{r}
risk_estimates(missl_risk_roll,
               risk_measures = c("VaR"),
               alpha = 0.01,
               exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est), col = yes_no_cols[1]) +
  geom_point(aes(x = row_num, y = realized), col = yes_no_cols[2], 
             inherit.aes = FALSE, data = . %>% filter(exceeded)) +
  labs(x = "estimation window",
       y = "portfolio log returns",
       col = "Exceeded",
       subtitle = paste0("Exceedances are highlighted in ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**red**</span>",
                     "."),
       title = "Risk measure: VaR with alpha level 0.01"
       ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")

risk_estimates(missl_risk_roll,
               risk_measures = c("ES_mean"),
               alpha = 0.01,
               exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est), col = yes_no_cols[1]) +
  geom_point(aes(x = row_num, y = realized), col = yes_no_cols[2], 
             inherit.aes = FALSE, data = . %>% filter(exceeded)) +
  labs(x = "estimation window",
       y = "portfolio log returns",
       col = "Exceeded",
       subtitle = paste0("Exceedances are highlighted in ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**red**</span>",
                     "."),
       title = "Risk measure: ES (mean estimation) with alpha level 0.01"
       ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")
```

### Conditional `missl` portfolio

Estimation is performed in the same fashion as above with some additional arguments concerning the conditioning on the stock Iberdrola.

```{r}
missl_cond_marginal_settings <- marginal_settings(
  train_size = 1000,
  refit_size = 50
)

missl_cond_vine_settings <- vine_settings(
  train_size = 200,
  refit_size = 25,
  family_set = "parametric",
  vine_type = "dvine"
)

future::plan("multisession", workers = 6)
missl_cond_risk_roll <- estimate_risk_roll(
  msci_spain_complete_data %>% select(-c(date, msci_spain_index)),
  weights = NULL,
  marginal_settings = missl_cond_marginal_settings,
  vine_settings = missl_cond_vine_settings,
  alpha = c(0.01, 0.05, 0.1),
  risk_measures = c("VaR", "ES_mean", "ES_median", "ES_mc"),
  n_samples = 1000,
  n_mc_samples = 1000,
  cond_vars = "iberdrola",
  cond_alpha = c(0.05, 0.5),
  trace = TRUE
)
future::plan("sequential")
```

```{r}
summary(missl_cond_risk_roll)
```

```{r}
# what about the evolution of the conditional values (iberdrola)
risk_estimates(missl_cond_risk_roll, risk_measures = "ES_mean",
               alpha = 0.05, exceeded = TRUE) %>%
  ggplot() +
  geom_line(
    data = msci_spain_complete_data[1001:nrow(msci_spain_complete_data), ],
    aes(x = 1001:nrow(msci_spain_complete_data),
        y = iberdrola), col = "lightgrey", size = .3) +
  geom_line(aes(x = row_num, y = iberdrola, col = factor(cond_alpha)),
            size = 0.5) +
  scale_color_manual(values = c(yes_no_cols[1], "#477042"),
                     name = "Conditional alpha level") +
  labs(title = "Conditional variable: Iberdrola",
       x = "estimation window", y = "log returns") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

Nice to see so we expect the corresponding conditional risk measures with low conditional $\alpha$ level to be way more conservative under the assumption that Iberdrola has a generally positive dependence with the portfolio of interest. Therefore look at the next plot.

```{r}
risk_estimates(missl_cond_risk_roll, risk_measures = "ES_mean",
               alpha = 0.05, exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est,
                col = factor(cond_alpha), linetype = factor(cond_alpha)),
            size = 0.5) +
  labs(x = "estimation window", y = "portfolio log returns",
       linetype = "Conditional alpha level",
       title = "Comparison of risk measure behaivior for 2 conditional alpha levels") +
  scale_color_manual(values = c(yes_no_cols[1], "#477042"),
                     name = "Conditional alpha level") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```


A close look wrt exceedences.

```{r}
risk_estimates(missl_cond_risk_roll, risk_measures = "ES_mean",
               cond_alpha = 0.5,
               alpha = 0.05, exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est), col = yes_no_cols[1]) +
  geom_point(aes(x = row_num, y = realized), col = yes_no_cols[2], 
             inherit.aes = FALSE, data = . %>% filter(exceeded)) +
  labs(x = "estimation window",
       y = "portfolio log returns",
       col = "Exceeded",
       subtitle = paste0("Exceedances are highlighted in ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**red**</span>",
                     "."),
       title = "Risk measure: ES (mean estimation) with alpha level 0.05, conditional alpha level 0.5"
       ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")

risk_estimates(missl_cond_risk_roll, risk_measures = "ES_mean",
               cond_alpha = 0.05,
               alpha = 0.05, exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est), col = yes_no_cols[1]) +
  geom_point(aes(x = row_num, y = realized), col = yes_no_cols[2], 
             inherit.aes = FALSE, data = . %>% filter(exceeded)) +
  labs(x = "estimation window",
       y = "portfolio log returns",
       col = "Exceeded",
       subtitle = paste0("Exceedances are highlighted in ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**red**</span>",
                     "."),
       title = "Risk measure: ES (mean estimation) with alpha and conditional alpha level 0.05"
       ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")
```

While the 0.5 conditional value leads to lots of exceedances (not surprisingly) the conditional value of 0.05 shows not many exceedances. For a rough comparison one can have a short comparison with the above fitted unconditional ES with mean estimation for the same confidence level.

```{r}
risk_estimates(missl_risk_roll,
               risk_measures = c("ES_mean"),
               alpha = 0.05,
               exceeded = TRUE) %>%
  ggplot() +
  geom_line(aes(x = row_num, y = realized), col = "lightgrey") +
  geom_line(aes(x = row_num, y = risk_est), col = yes_no_cols[1]) +
  geom_point(aes(x = row_num, y = realized), col = yes_no_cols[2], 
             inherit.aes = FALSE, data = . %>% filter(exceeded)) +
  labs(x = "estimation window",
       y = "portfolio log returns",
       col = "Exceeded",
       subtitle = paste0("Exceedances are highlighted in ",
                     "<span style='color:",
                     yes_no_cols[2],
                     "'>**red**</span>",
                     "."),
       title = "Risk measure: ES (mean estimation) with alpha level 0.05"
       ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")
```

So in the direct comparison the conditional risk measure is way more conservative and might help in stresstesting situations for projected hard times of specific conditional assets.
